Static Analysis by Path Focusing

### 2011-05-06 report/report.tex
report

\documentclass[a4paper,english,titlepage,11pt]{article}

@@

\part{Introduction}

@@

\part{Abstract Interpretation: state of the art}
\label{stateoftheartpart}

@@

For instance, considering the program:
\begin{C}
for (int i=0; i < 100; i++) {
}
\end{C}
The analysis starts with $i=0$. After one iteration, one obtain $i \in
[0,1]$, \dots. After a widening operation, the result becomes $i \in [0,
+\infty]$, which is an invariant, but very imprecise. We can do a new iteration,
with $i\in [0, +\infty]$ at the beginning of the loop, and we find that the
values for $i$ at the next step are in $[0, 100]$. This is still an invariant,
but much better.

@@

\part{Path Focusing technique} \label{pathfocusingpart}

@@

	The essence of the method is to
	choose an integer $\delta \in \{1,..,m\}$, and a mapping function
	$\sigma: \{1,..,m\} \times \{1,..,n\} \mapsto \{1,..,m\}$.
	For each polyhedra of the disjunctive invariant, and for each path in the
	graph, the image of the polyhedra $X_{p,j}$ by the transition
	$\tau_{p,i}$ is joined with $X_{p,\sigma(j,i)}$.

@@

\FloatBarrier
\part{Implementation}\label{implementationpart}

@@

\begin{figure}[!h]
\begin{minipage}[c]{.45\linewidth}
\begin{C}
int main() {
   int x = 0;
   int y = 0;

@@

      if (y < 0) break;
      x++;
   }
}
\end{C}
\end{minipage}
\begin{minipage}[c]{.45\linewidth}
\includegraphics[width=8.5cm]{"images/cfg_gopan"}
\end{minipage}
\caption{Our running example and its internal representation in SSA form}
\label{IR}
\end{figure}
\FloatBarrier

As we see if Figure \ref{IR}, LLVM IR is slightly different from
our definition of control flow graph: it uses the notion of \emph{BasicBlock}.
A \emph{BasicBlock} is a series of instructions that are necessarily executed
one after the others, meaning that if the first intruction of a BasicBlock is
executed, then the rest of the instructions are necessarily executed exactly
once, in order.

@@

This is the case for the distinction between \emph{signed} and \emph{unsigned}
integers, which is inexistant in the representation, since the compiler doesn't
need this information to generate code. Yet, for static analysis, one need to
have this distinction to improve precision:
for instance, if $n$ is unsigned, one could add the constraint $n \geq 0$ in our
	abstract value at the beginning of the analysis.
	This lack of precision may be recovered by using disjunctive invariants: one
	could attach two polyhedra to the control point: the first one with the
	constraint $n \geq 0$, and the second one with $n < 0$.

@@

	We attach only one abstract value to each BasicBlock of the IR. Indeed,
	there would be a huge memory consumption if we would attach a polyhedra
	after each instruction. A BasicBlock begins with all its $\Phi$-variables,
	and ends with the other instructions. We choose to attach our abstract value
	just after the $\Phi$-variables assignments. In this way, the
	$\Phi$-variables of the associate BasicBlock will be part of the dimensions
	of the polyhedron.

@@

\part{Future Work}
\label{future}
\emph{Path Focusing} technique allows to work on control flow graphs with a huge
number of paths, since they are computed only when needed. With this in mind,
one could think of several improvements using path focusing, by encoding into
the control flow graph new informations as paths, such as arithmetic overflows,
alias analysis, \dots
In this part, we explain how to encode such kind of informations into implicit
paths of the CFG.

@@

\part{Conclusion}

>>>>

\documentclass[a4paper,english,titlepage,11pt]{report}

@@

\chapter{Introduction}

@@

\chapter{Abstract Interpretation: state of the art}
\label{stateoftheartpart}

@@

For instance, considering the program:
\begin{C}
for (int i=0; i < 100; i++) {
}
\end{C}
The analysis starts with $i=0$. After one iteration, one obtain $i \in
[0,1]$, \dots. After a widening operation, the result becomes $i \in [0,
+\infty]$, which is an invariant, but very imprecise. We can do a new iteration,
with $i\in [0, +\infty]$ at the beginning of the loop, and we find that the
values for $i$ at the next step are in $[0, 100]$. This is still an invariant,
but much better. We can repeat such iterations as long as the set narrows, and
we always obtain a correct over-approximation.
In practice, a single narrowing iteration is enough to recover the precision.

@@

\chapter{Path Focusing technique} \label{pathfocusingpart}

@@

	The principle of the method is to
	choose an integer $\delta \in \{1,..,m\}$, and a mapping function
	$\sigma: \{1,..,m\} \times \{1,..,n\} \mapsto \{1,..,m\}$.
	For each polyhedra of the disjunctive invariant, and for each path in the
	graph, the image of the polyhedra $X_{p,j}$ by the transition
	$\tau_{p,i}$ is joined with $X_{p,\sigma(j,i)}$.

@@

\FloatBarrier
\chapter{Implementation}\label{implementationpart}

@@

\begin{figure}[!h]
\begin{minipage}[c]{.35\linewidth}
{\small
\begin{C}
int main() {
   int x = 0;
   int y = 0;

@@

      if (y < 0) break;
      x++;
   }
}
\end{C}
}
\end{minipage}
\begin{minipage}[c]{.69\linewidth}
%\includegraphics[width=8.5cm]{"images/cfg_gopan"}

{\tiny
% \begin{tikzpicture}[anchor=mid,>=latex',line join=bevel,]
\begin{tikzpicture}[>=latex',line join=bevel,scale=0.5]
  \pgfsetlinewidth{1bp}
%%
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (273bp,13bp) node {CFG for 'main' function};
\end{scope}
  \pgfsetcolor{black}
  % Edge: Node0x9828da8 -> Node0x9828dd8
  \draw [->] (356.55bp,400.82bp) .. controls (333.72bp,390.36bp) and (308.41bp,378.75bp)  .. (274.21bp,363.07bp);
  % Edge: Node0x9828b18 -> Node0x9828d48
  \draw [->] (310bp,694.72bp) .. controls (310bp,686.41bp) and (310bp,677bp)  .. (310bp,657.09bp);
  % Edge: Node0x9828dd8 -> Node0x9828e08
  \draw [->] (70bp,241bp) .. controls (70bp,229.11bp) and (70bp,216.19bp)  .. (70bp,194.35bp);
  % Edge: Node0x9828d48 -> Node0x9828da8
  \draw [->] (456bp,530bp) .. controls (475.15bp,530bp) and (473.05bp,509.87bp)  .. (461.87bp,479.21bp);
  % Edge: Node0x9828e08 -> Node0x9828e68
  \draw [->] (70bp,133.91bp) .. controls (70bp,122.63bp) and (70bp,109.61bp)  .. (70bp,87.235bp);
  % Edge: Node0x9828d78 -> Node0x9828dd8
  \draw [->] (155.32bp,400.82bp) .. controls (154.06bp,392.13bp) and (152.69bp,382.65bp)  .. (149.87bp,363.2bp);
  % Edge: Node0x9828d48 -> Node0x9828d78
  \draw [->] (164bp,530bp) .. controls (145.24bp,530bp) and (143.41bp,510.23bp)  .. (148.69bp,479.21bp);
  % Edge: Node0x9828e38 -> Node0x9828d48
  \draw [->] (285.99bp,203.34bp) .. controls (287.87bp,215.1bp) and (289.73bp,228.04bp)  .. (291bp,240bp) .. controls (300.66bp,331.07bp) and (305.54bp,436.13bp)  .. (308.25bp,516.72bp);
  % Edge: Node0x9828dd8 -> Node0x9828e38
  %\draw [->] (284bp,254bp) .. controls (293.39bp,254bp) and (292.88bp,234.23bp)  .. (287.99bp,203.21bp);
  \draw [->] (210bp,240bp) -- (240.99bp,203.21bp);
  % Node: Node0x9828d48
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (165bp,517bp) -- (165bp,657bp) -- (455bp,657bp) -- (455bp,517bp) -- cycle;
  \draw (165bp,543bp) -- (455bp,543bp);
  \draw (310bp,517bp) -- (310bp,543bp);
  \draw (173bp,639bp) node[right] {\%1:};
  \draw (173bp,605bp) node[right] { \%x.0 = phi i32 [ 0, \%0 ], [ \%11, \%10 ]};
  \draw (173bp,587bp) node[right] { \%y.1 = phi i32 [ 0, \%0 ], [ \%y.0, \%10 ]};
  \draw (173bp,569bp) node[right] { \%2 = icmp sle i32 \%x.0, 50};
  \draw (173bp,551bp) node[right] { br i1 \%2, label \%3, label \%5};
  \draw (238bp,525bp) node {T};
  \draw (383bp,525bp) node {F};
\end{scope}
  % Node: Node0x9828b18
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (261bp,695bp) -- (261bp,755bp) -- (360bp,755bp) -- (360bp,695bp) -- cycle;
  \draw (269bp,737bp) node[right] {\%0:};
  \draw (269bp,703bp) node[right] { br label \%1};
\end{scope}
  % Node: Node0x9828e68
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (30bp,27bp) -- (30bp,87bp) -- (111bp,87bp) -- (111bp,27bp) -- cycle;
  \draw (38bp,69bp) node[right] {\%12:};
  \draw (38bp,35bp) node[right] { ret i32 0};
\end{scope}
  % Node: Node0x9828e38
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (172bp,125bp) -- (172bp,203bp) -- (387bp,203bp) -- (387bp,125bp) -- cycle;
  \draw (180bp,185bp) node[right] {\%10:};
  \draw (180bp,151bp) node[right] { \%11 = add nsw i32 \%x.0, 1};
  \draw (180bp,133bp) node[right] { br label \%1};
\end{scope}
  % Node: Node0x9828da8
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (338bp,401bp) -- (338bp,479bp) -- (547bp,479bp) -- (547bp,401bp) -- cycle;
  \draw (346bp,461bp) node[right] {\%5:};
  \draw (346bp,427bp) node[right] { \%6 = add nsw i32 \%y.1, -1};
  \draw (346bp,409bp) node[right] { br label \%7};
\end{scope}
  % Node: Node0x9828d78
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (59bp,401bp) -- (59bp,479bp) -- (263bp,479bp) -- (263bp,401bp) -- cycle;
  \draw (67bp,461bp) node[right] {\%3:};
  \draw (67bp,427bp) node[right] { \%4 = add nsw i32 \%y.1, 1};
  \draw (67bp,409bp) node[right] { br label \%7};
\end{scope}
  % Node: Node0x9828dd8
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (-1bp,241bp) -- (-1bp,363bp) -- (283bp,363bp) -- (283bp,241bp) -- cycle;
  \draw (-1bp,267bp) -- (283bp,267bp);
  \draw (142bp,241bp) -- (142bp,267bp);
  \draw (8bp,345bp) node[right] {\%7:};
  \draw (8bp,311bp) node[right] { \%y.0 = phi i32 [ \%4, \%3 ], [ \%6, \%5 ]};
  \draw (8bp,293bp) node[right] { \%8 = icmp slt i32 \%y.0, 0};
  \draw (8bp,275bp) node[right] { br i1 \%8, label \%9, label \%10};
  \draw (71bp,249bp) node {T};
  \draw (212bp,249bp) node {F};
\end{scope}
  % Node: Node0x9828e08
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (16bp,134bp) -- (16bp,194bp) -- (124bp,194bp) -- (124bp,134bp) -- cycle;
  \draw (24bp,176bp) node[right] {\%9:};
  \draw (24bp,142bp) node[right] { br label \%12};
\end{scope}
%
}

\end{minipage}
\caption{Our running example and its internal representation in SSA form}
\label{IR}
\end{figure}
\FloatBarrier

As we see if Figure \ref{IR}, LLVM IR is slightly different from
our definition of control flow graph: it uses the notion of \emph{BasicBlock}.
A \emph{BasicBlock} is a sequence of instructions that are necessarily executed
one after the others, meaning that if the first intruction of a BasicBlock is
executed, then the rest of the instructions are necessarily executed exactly
once, in order.

@@

This is the case for the distinction between \emph{signed} and \emph{unsigned}
integers, which is inexistant in the representation, since the compiler doesn't
need this information to generate code. Yet, for static analysis, one need to
have this distinction to improve precision:
for instance, if $n$ is unsigned, one could add the constraint $n \geq 0$ in our
	abstract value at the beginning of the analysis.
	This lack of precision may be limited by using disjunctive invariants: one
	could attach two polyhedra to the control point: the first one with the
	constraint $n \geq 0$, and the second one with $n < 0$.

@@

	We attach only one abstract value to each BasicBlock of the IR. Indeed,
	there would be a huge memory consumption if we would attach a polyhedron
	after each instruction. A BasicBlock begins with all its $\Phi$-variables,
	and ends with the other instructions. We choose to attach our abstract value
	just after the $\Phi$-variables assignments. In this way, the
	$\Phi$-variables of the associate BasicBlock will be part of the dimensions
	of the polyhedron.

@@

\chapter{Future Work}
\label{future}
\emph{Path Focusing} technique allows to work on control flow graphs with a huge
number of paths, since they are computed only when needed. With this in mind,
one could think of several improvements using path focusing, by encoding into
the control flow graph new informations as paths, such as arithmetic overflows,
alias analysis, \dots
In this part, we explain how to encode such kind of informations into implicit
paths of the CFG.

@@

	\section{Combining Lookahead Widening and Path Focusing}

	Path Focusing technique works on the expanded multigraph.
	Its limitation is that it does narrowing
	sequences during the iterations only for self-loops of this multigraph.

	Lookahead Widening works on the classical control flow graph, and stabilizes
	temporarily the fixpoint analysis on a part of this graph, forgetting paths
	that are not feasible at the first iteration.

	One could try to combine these two techniques, and apply Lookahead Widening
	on the expanded multigraph, where the different paths are maintained
	implicit thanks to Path Focusing technique. The advantage of this approach
	compared to the classical Path Focusing is that narrowing iterations could
	be computed on loops that are not self-loops.

\chapter{Conclusion}
